\documentclass[10pt]{article}
\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{tcolorbox}
\usepackage{booktabs}

\linespread{1.08}
\setlength{\parskip}{0.4ex plus 0.2ex minus 0.1ex}

\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titlespacing*{\section}{0pt}{1.5ex}{1ex}
\titlespacing*{\subsection}{0pt}{1ex}{0.5ex}

\setlist{itemsep=1pt, topsep=3pt, parsep=1pt, leftmargin=1.5em}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small EECS 127}
\fancyhead[R]{\small Discussion 8}
\fancyfoot[C]{\small\thepage}
\renewcommand{\headrulewidth}{0.4pt}

\newcommand{\RR}{\mathbb{R}}
\newcommand{\NN}{\mathcal{N}}
\newcommand{\Range}{\mathcal{R}}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\trace}{tr}

\begin{document}

\begin{center}
{\large\bfseries EECS 127/227AT \hfill Optimization Models in Engineering \hfill UC Berkeley}\\[4pt]
{\Large\bfseries Discussion 8} \hfill {\large Spring 2026}
\end{center}

\vspace{1em}

\section{Convexity and Composition of Functions}

Let $f : \RR \to \RR$ and $g : \RR^n \to \RR$. Define the composition of $f$ with $g$ as $h = f \circ g : \RR^n \to \RR$ such that
$h(\vec{x}) = f(g(\vec{x}))$.

\begin{enumerate}[label=(\alph*)]
    \item Show that if $f$ is convex and non decreasing and $g$ is convex, then $h$ is convex.

    \vspace{4cm}

    \item Show that there exists $f$ non decreasing and $g$ convex, such that $h = f \circ g$ is not convex.

    \vspace{4cm}

    \item Show that there exists $f$ convex and $g$ convex such that $h = f \circ g$ is not convex.
\end{enumerate}

\newpage

\section{Simple Constrained Optimization Problem}

Consider the optimization problem
\begin{alignat}{3}
    &\min_{x_1, x_2 \in \RR} &\quad& f(x_1, x_2) \\
    &\text{s.t.} && 2x_1 + x_2 \geq 1 \\
    &&& x_1 + 3x_2 \geq 1 \\
    &&& x_1 \geq 0,\; x_2 \geq 0
\end{alignat}

\begin{enumerate}[label=(\alph*)]
    \item Make a sketch of the feasible set.

    \vspace{4cm}
\end{enumerate}

For each of the following objective functions, give the optimal set or the optimal value.

\begin{enumerate}[label=(\alph*), resume]
    \item $f(x_1, x_2) = x_1 + x_2$.

    \vspace{3cm}

    \item $f(x_1, x_2) = -x_1 - x_2$.

    \vspace{3cm}

    \item $f(x_1, x_2) = x_1$

    \vspace{3cm}

    \item $f(x_1, x_2) = \max\{x_1, x_2\}$

    \vspace{3cm}

    \item $f(x_1, x_2) = x_1^2 + 9x_2^2$
\end{enumerate}

\vfill
\noindent\textcopyright{} UCB EECS 127/227AT, Spring 2026. All Rights Reserved. This may not be publicly shared without explicit permission.

\end{document}
